---
title: "625-final"
author: "Guanyi Wu"
date: "2025-12-03"
output: pdf_document
---

Dataset exploration

pre-load

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(kableExtra)
```

load dataset

```{r}
dats <- read.csv("StudentPerformancePredict/data/raw/StudentPerformanceFactors.csv")
```

shape and type

```{r}
shape <- dim(dats)
summary(dats)
```

7 numeric columns: (1)Hours_Studied, (2)Attendance, (3)Sleep_Hours, (4)Previous_Scores, (5)Tutoring_Sessions, (6)Physical_Activity, (7)Exam_Score

13 categorical columns: (1)Parental_Involvement, (2)Access_to_Resources, (3)Extracurricular_Activities, (4)Motivation_Level, (5)Internet_Access, (6)Family_Income, (7)Teacher_Quality, (8)School_Type, (9)Peer_Influence, (10)Learning_Disabilities, (11)Parental_Education_Level, (12)Distance_from_Home, (13)Gender

Missing value

```{r}
dats[dats == ""] <- NA
colSums(is.na(dats))
na_df <- data.frame(
  variable = names(dats),
  na_count = colSums(is.na(dats))
)

ggplot(na_df, aes(x = variable, y = na_count)) +
  geom_col() +
  coord_flip() +
  labs(title = "Missing Values per Column",
       x = "Column",
       y = "Number of NA") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))
```

Check the distuibution of columns with missing value

```{r}
cols <- c("Parental_Education_Level",
          "Teacher_Quality",
          "Distance_from_Home")

for (col in cols) {
  
  df_plot <- dats %>%
    filter(!is.na(.data[[col]])) %>%
    count(.data[[col]]) %>%
    arrange(desc(n))
  
  p <- ggplot(df_plot, aes(x = .data[[col]], y = n, fill = .data[[col]])) +
  geom_col(width = 0.7, alpha = 0.9, color = "white") +
  geom_text(aes(label = n),
            vjust = -0.3,
            size = 5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = col, x = "", y = "Count") +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  expand_limits(y = max(df_plot$n) * 1.15)
  
  print(p)
}
```

Replace the missing values in each column with its most frequent value.

```{r}
dats <- dats %>%
  mutate(
    Parental_Education_Level = replace_na(Parental_Education_Level, "High School"),
    Teacher_Quality          = replace_na(Teacher_Quality, "Medium"),
    Distance_from_Home       = replace_na(Distance_from_Home, "Near")
  )
#colSums(is.na(dats))
```

Also, we found outlier value 101 in Exam Score, so we replace it to 100.

```{r}
dats$Exam_Score[dats$Exam_Score == 101] <- 100
summary(dats)
```

Next we check the distribution of each column

for numeric columns

```{r}
numeric_cols <- c(
  "Hours_Studied", "Attendance", "Sleep_Hours",
  "Previous_Scores", "Tutoring_Sessions",
  "Physical_Activity", "Exam_Score"
)


num_colors <- RColorBrewer::brewer.pal(n = 7, "Set2")

dats %>%
  select(all_of(numeric_cols)) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "value") %>%
  mutate(variable = factor(variable, levels = numeric_cols)) %>%
  ggplot(aes(x = value, fill = variable)) +
  geom_density(alpha = 0.6, color = NA) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = num_colors) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Distribution of Numerical Features",
    x = "", y = ""
  )
```

for categorical columns

```{r,fig.width = 12, fig.height = 20}
categorical_cols <- c(
  "Parental_Involvement", "Access_to_Resources",
  "Extracurricular_Activities", "Motivation_Level",
  "Internet_Access", "Family_Income",
  "Teacher_Quality", "School_Type",
  "Peer_Influence", "Learning_Disabilities",
  "Parental_Education_Level", "Distance_from_Home",
  "Gender"
)

cat_colors <- RColorBrewer::brewer.pal(n = 12, "Set3")
cat_colors <- c(cat_colors, "#A8D5BA")  

dats %>%
  select(all_of(categorical_cols)) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "value") %>%
  mutate(variable = factor(variable, levels = categorical_cols)) %>%
  ggplot(aes(x = value, fill = variable)) +
  geom_bar(color = "white") +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.2, size = 3
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  scale_fill_manual(values = cat_colors) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1),
    strip.text = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Distribution of Categorical Features",
    x = "", y = ""
  )
```

```{r}
num_dat <- dats %>% select(where(is.numeric))

summary_tbl <- data.frame(
  Variable = names(num_dat),
  Min      = sapply(num_dat, min,    na.rm = TRUE),
  Q1       = sapply(num_dat, quantile, 0.25, na.rm = TRUE),
  Median   = sapply(num_dat, median, na.rm = TRUE),
  Mean     = sapply(num_dat, mean,   na.rm = TRUE),
  Q3       = sapply(num_dat, quantile, 0.75, na.rm = TRUE),
  Max      = sapply(num_dat, max,    na.rm = TRUE)
)

summary_tbl %>%
  kable(digits = 2, caption = "Summary statistics for numeric variables") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


corralation
```{r}
library(ggplot2)
library(reshape2)


X <- model.matrix(~ . - 1, data = dats)
cor_mat <- cor(X, use = "pairwise.complete.obs")


cm_long <- reshape2::melt(cor_mat, varnames = c("Var1", "Var2"), value.name = "corr")


cm_long$Var1 <- factor(cm_long$Var1, levels = rownames(cor_mat))
cm_long$Var2 <- factor(cm_long$Var2, levels = colnames(cor_mat))


target_col <- tail(colnames(cor_mat), 1)

ggplot(cm_long, aes(x = Var2, y = Var1, fill = corr)) +

  geom_tile(aes(color = Var2 == target_col), linewidth = 0.7) +
  scale_color_manual(values = c("FALSE" = NA, "TRUE" = "yellow"), guide = "none") +


  geom_text(aes(label = sprintf("%.1f", corr)), size = 2) +

  
  scale_fill_gradient2(
    low = "#313695",
    mid = "white",
    high = "#A50026",
    midpoint = 0,
    limits = c(-1, 1)
  ) +

  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 8) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid = element_blank()
  )
```

Linear relationship

```{r}
library(viridis)
ggplot(dats, aes(x = Hours_Studied,
                 y = Exam_Score,
                 color = Attendance)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis(
    option = "D",
    name   = "Attendance"
  ) +
  labs(
    title = "Study Hours and Attendance on Exam",
    x = "Hours_Studied",
    y = "Exam_Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```












